The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-02-07 Paper.Rmd
trivial
minor edits to the Q2 sectios

Let $E_i(t)$ be an ecological variable 
measured in location $i$ ($i=1,2$).
We initially assumed the variable follows AR(1) dynamics within a location, 
$E_i(t+1)=\beta E_{i}(t-1)+ \sqrt{1-\beta^2} \epsilon_i(t)$. 
We used $\beta=0.5$. Here the $\epsilon_i(t)$ represent environmental noise; they are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The ecological variable, $E$, is very general, and could represent, for instance, deviations
of a population density from a carrying capacity or average density, deviations of
total plant community biomass or other community variable from an average value,
or flux of a biogeochemical variable.

To explore the importance of nonlinear dynamics for our hypothersis, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases. The $\epsilon_i(t)$ were again standard
normal and independent through time, but were related (see below) across locations.

@@

The second mechanism we will explore, which was also summarized in the 
Introduction,
relates to Liebig's law of the minimum, or, more generally, to the fact that 
environmental variables can often influence ecological variables through nonlinear 
relationships. If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored using a model 
whether the ecological variable
could then exhibit asymmetric tail dependence between measurements made in two
locations. Let $E_i(t)$ again be an ecological variable measured in locations
$i=1,2$. Assume $E_i(t)$ has dynamics that follow 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where $g$ is a function
describing how the environmental variable $\epsilon_i(t)$ influences
$E_i(t)$, and the $\delta_i(t)$ are standard-normally distributed 
local noise, independent across time and sampling locations. We used $b=0.1$ and
$a=0.2$, and $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. We used $g$ equal to $g_1$ or $g_2$ in two different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise; and 
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represent environmental effects that occur only for negative 
values of the environmental variables, and $g_2$ represent environmental effects 
that occur only for positive values. Note that both $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, being drawn from 
bivariate normal distributions. Hence the Moran-like mechanism discussed
previously cannot be operating in this model.

We simulated the model once using $g_1$ and once using $g_2$, for 5000 time
steps in both cases, retaining $E_1(t)$ and $E_2(t)$ for the final 500 time 
steps of each simulation. We applied our model selection algorithms and 
non-parametric statistics, which were previously applied to 
data to answer Q1, to the 
simulation outputs to discover if the model could
give rise to non-N copula structure and asymmetric tail dependence in the 
ecological variable.

@@

For weak noise and using our nonlinear model (methods for Q2), copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_0.1}).
Though there were again significant but small differences for some statistics
and simulations, the overall hypothesis that copula structure in the noise translates
to similar but not identical copula structure in output populations, at least
as measured with the tail dependence and correlation statistics we used, was
supported. Since many ecological models are nonlinear, this result provides the 
important expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure for typical ecological dynamics,
as long as the noise is small enough that dynamics stay relatively close
to model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=X$, with similar results. The deterministic 
<!--***DAN: Shya, please replace X in the above line with the actual value(s)
of r that you used.-->
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (so-called under-compensating dynamics), but exhibits 
an oscillatory approach to a stable equilibrium at $K$ when $r>1$ (over-compensating
dynamics).

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise results, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs; and likewise for SG and SJ copulas. Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=X$, with similar results.
<!--***DAN: Right, Shya? Please replace X with the actual value(s)-->

Our model with environmental effects which differed in the
two tails produced simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure (Fig. \ref{fig_ms_asym_b_0.1}).
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) and
our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) 
statistically confirmed the visual result.
We repeated our analyses with $b=-0.1, 0.9, -0.9$, with similar conclusions
in all cases.
<!--***DAN: Right Shya? I have written this, but I want you to show me these results
(without putting them in the paper) to make sure the statement is accurate.
But we should put the figs in the sup mat, not the stats results, I think, but we could put the figs. Or maybe we do want to put the stats. Let's look at them anyway. Maybe the sentence needs to
be rewritten if the results for other b are actually different.-->

\begin{figure}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/asym_sens_results/a_0.2_b_0.1_r_0.8_asym_sens.pdf}
\caption{If environmental effects operate asymmetrically in their tails on
ecological variables, it can result in non-N copula structure and asymmetric 
tail dependence across space in the ecological variables. Shown are the last
$500$ points from simulations described in the text.\label{fig_ms_asym_b_0.1}}
\end{center}
\end{figure}
<!--***DAN: Shya, pls remove the left 4 panels, and also the headers of the right
two panels - we can put that information in the caption, as I have done-->

>>>>

Let $E_i(t)$ be an ecological variable 
measured in location $i$ ($i=1,2$).
We initially assumed the variable follows AR(1) dynamics within a location, 
$E_i(t+1)=\beta E_{i}(t-1)+ \sqrt{1-\beta^2} \epsilon_i(t)$. 
We used $\beta=0.5$. Here the $\epsilon_i(t)$ represent environmental noise; they are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The ecological variable, $E$, is very general, and could represent, for instance, deviations
of a population density from a carrying capacity or average density, deviations of
total plant community biomass or other community variable from an average value,
or flux of a biogeochemical variable such as methane or CO$_2$..

To explore the importance of nonlinear dynamics for our hypothesis, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model strongly into play. 
The $\epsilon_i(t)$ were again standard
normal and independent through time, but were related (see below) across locations.

@@

The second mechanism we explored, which was also summarized in the 
Introduction,
relates to Liebig's law of the minimum, or, more generally, to the fact that 
environmental variables can often influence ecological variables through nonlinear 
relationships. If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored using a model 
whether the ecological variable
could then exhibit asymmetric tail dependence between measurements made in two
locations. Let $E_i(t)$ again be an ecological variable measured in locations
$i=1,2$. Assume $E_i(t)$ has dynamics that follow 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where $g$ is a function
describing how the environmental variable $\epsilon_i(t)$ influences
$E_i(t)$, and the $\delta_i(t)$ are standard-normally distributed 
local noise, independent across time and sampling locations. We used $b=0.1, -0.1, 0.9, -0.9$ and
$a=0.2$, and $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. Negative values of $b$ correspond to overcompensating dynamics 
and positive values correspond to undercompensating dynamics; larger-magnitude $b$
mean slower return time of the model to equilibrium after a perturbation. 
We used $g$ equal to $g_1$ or $g_2$ in two different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise; and 
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represent environmental effects that occur only for negative 
values of the environmental variables, and $g_2$ represent environmental effects 
that occur only for positive values. Note that both $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, being drawn from 
bivariate normal distributions. Hence the Moran-like mechanism discussed
previously cannot be operating in this model.

For each value of $b$, we simulated the model once using $g_1$ and once using $g_2$, for 5000 time
steps in both cases, retaining $E_1(t)$ and $E_2(t)$ for the final 500 time 
steps of each simulation. We applied our model selection algorithms and 
non-parametric statistics, which were previously applied to 
data to answer Q1, to the 
simulation outputs to discover if the model could
give rise to non-N copula structure and asymmetric tail dependence in the 
ecological variable.

@@

For weak noise and using our nonlinear model (methods for Q2), copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_0.1}).
Though there were again significant but small differences for some statistics
and simulations, the overall hypothesis that copula structure in the noise translates
to similar but not identical copula structure in output populations, at least
as measured with the tail dependence and correlation statistics we used, was
supported. Since many ecological models are nonlinear, this result provides the 
important expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure for typical ecological dynamics,
as long as the noise is small enough that dynamics stay relatively close
to the model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=X$, with similar results. The deterministic 
<!--***DAN: Shya, please replace X in the above line with the actual value(s)
of r that you used. But please be sure
the results were "similar". I encourage you to show me the actual results if there is 
any doubt in your mind.-->
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (undercompensating dynamics), but exhibits 
an oscillatory approach to $K$ when $r>1$ (overcompensating
dynamics).

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise cases, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs; and likewise for SG and SJ copulas. Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=X$, with similar results.
<!--***DAN: Right, Shya? Please replace X with the actual value(s), but please be sure
the results were "similar". I encourage you to show me the actual results if there is 
any doubt in your mind.-->

Our model with environmental effects which differed in the
two tails produced simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure (Fig. \ref{fig_ms_asym_b_0.1}).
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) and
our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) 
statistically confirmed the visual result.
We repeated our analyses with $b=-0.1, 0.9, -0.9$, with similar conclusions
in all cases.
<!--***DAN: Dan to rewrite this para after seeing the complete collection of results-->

\begin{figure}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/asym_sens_results/a_0.2_b_0.1_r_0.8_asym_sens.pdf}
\caption{If environmental effects operate asymmetrically in their tails on
ecological variables, it can result in non-N copula structure and asymmetric 
tail dependence across space in the ecological variables. Shown are the last
$500$ points from simulations described in the text.\label{fig_ms_asym_b_0.1}}
\end{center}
\end{figure}
